Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
references

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response. First, tree height may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water. Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.

@@

Consistent with studies in other forests worldwide (**REFS**), taller trees in this forest exhibited lower drought resistance. 
Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand (**REFS**). 
Unlike other temperate forests where modestly cooler understory conditions have been documented (**REFS**), particularly under drier conditions (**REFS**), we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman REF*). 
Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.   
Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms.

*Saving this for references: Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017] .*

>>>>

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response. First, tree height may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand (**REFS--KAT**)[@kunert_revised_2017]. Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water. Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.

@@

Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance. 
Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
Unlike other temperate forests where modestly cooler understory conditions have been documented (*Zellweger et al. 2019*), particularly under drier conditions (*Davis et al. 2019*), we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman; Rey-Sanchez et al. 2016*). 
Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.   
Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees fo different size.
